---
title: "Canadian Wind Turbines 💨"
output: github_document
editor_options: 
  chunk_output_type: console
---

# TidyTuesday 2020-10-27

If you want to join the next CorrelAid TidyTuesday Meetup, make sure to sign up to our [Newsletter](https://correlaid.us12.list-manage.com/subscribe?u=b294bf2834adf5d89bdd2dd5a&id=915f3f3eff) or reach out to us on [Twitter](https://twitter.com/CorrelAid)!

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

```

```{r Load, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidytuesdayR)
library(ragg)
library(sf)
#library(fontawesome) # for knitting, install with devtools::install_github("rstudio/fontawesome")

theme_set(theme_minimal(base_family = "Roboto Condensed"))

tt <- tt_load("2020-10-27")
df_wind <- tt$`wind-turbine`
```

```{r fig.width = 9, fig.height = 6}
ggplot(df_wind, aes(province_territory)) +
  geom_bar() +
  coord_flip()

df_wind %>% 
  count(province_territory) %>% 
  ggplot(aes(forcats::fct_reorder(province_territory, n), n)) +
  geom_col() +
  coord_flip()

df_wind %>% 
  count(project_name) %>% 
  filter(n > 50) %>% 
  ggplot(aes(forcats::fct_reorder(project_name, n), n)) +
  geom_col() +
  coord_flip()
```

```{r}
sf_wind <- 
  df_wind %>% 
  st_as_sf(coords = c("longitude", "latitude"), 
           crs = "+proj=longlat +datum=WGS84 +no_defs") %>% 
  st_transform(crs = "+proj=lcc +lon_0=-90 +lat_1=33 +lat_2=45")

sf_canada <-
  rnaturalearth::ne_countries(scale = 110, country = "Canada", 
                              returnclass = "sf") %>% 
  st_transform(crs = st_crs(sf_wind))

ggplot(sf_canada) +
  geom_sf(color = NA, fill = "grey92") +
  geom_sf(data = sf_wind, color = "#1D994E", alpha = .1, 
          size = 2, shape = 21, fill = NA)

ggplot(df_wind, aes(longitude, latitude)) +
  geom_hex()
```

**Tipps'n'Tricks:**

* `GGally::ggpairs()` for a quick EDA
* `rnaturalearth::ne_countries()` and `rnaturalearth::ne_states()` for shape files of countries
* `{ggforce}` package for many many cool things in ggplot2
* `forcats::fct_reorder()` and `forcats::fct_lump()` to reorder factors based on variables or merge those of low interest into a "other" class
* `theme(plot.title.position = "plot", plot.caption.position = "plot")` since `{ggplot2}` v3.0.0 to justify the title, subtitle and caption with the plot area not the panel border
* [Type sorted list of color palettes in R](https://github.com/EmilHvitfeldt/r-color-palettes/blob/master/type-sorted-palettes.md#qualitative-color-palettes)